---
title: "Creating a Metabolic ADLB ADaM"
output:
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating a Metabolic ADLB ADaM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(admiraldev)
```

# Introduction

This article describes creating a laboratory ADaM for metabolic clinical trials.

We advise you first consult the `{admiral}` [Creating a BDS Finding ADaM vignette](https://pharmaverse.github.io/admiral/articles/bds_finding.html).
The programming workflow around creating the general set-up of an `ADLB` using `{admiral}` functions is the same. 
In this vignette, we focus on common ADLB derivations in metabolic studies and avoid repeating information and maintaining the same content in two places. 

**Note**: *All examples assume CDISC SDTM and/or ADaM format as input unless otherwise specified.*

## Required Packages

The examples of this vignette require the following packages.

```{r, warning=FALSE, message=FALSE}
library(admiral)
library(admiralmetabolic)
library(pharmaversesdtm)
library(dplyr)
library(stringr)
```

# Programming Workflow

-   [Read in Data](#readdata)
-   [Define Lookup Tables](#lookups)
-   [Derive Core ADLB Variables](#derive_core)
-   [Derive Visit Information](#derive_visits)
-   [Derive Metabolic Parameters](#metabolic_params_desc)
    -   [Add variables required for derivation](#merge)
    -   [Derive HOMA-IR index](#derive_homa)
    -   [Derive FLI score](#derive_fli)
    -   [Derive FLI screening risk bands](#fli_bands)
-   [Remaining ADLB Set-up](#adlb_end)
-   [Example Script](#example)

## Read in Data {#readdata}

To start, all data frames needed for the creation of the ADaM dataset should be loaded into the global environment. Reading data will usually be a company specific process, however, for the purpose of this vignette, we will use example data from `{pharmaversesdtm}` and `{admiralmetabolic}`. We will utilize `LB` and `ADSL` data.

```{r, message=FALSE, warning=FALSE}
lb_metabolic <- pharmaversesdtm::lb_metabolic
admiralmetabolic_adsl <- admiralmetabolic::admiralmetabolic_adsl

lb <- convert_blanks_to_na(lb_metabolic)
adsl <- convert_blanks_to_na(admiralmetabolic_adsl)
```

## Define Lookup Tables {#lookups}

Define parameter lookup table used to derive `PARAMCD`, `PARAM`, and `PARAMN` variables. 

```{r}
# Assign PARAMCD, PARAM, and PARAMN
param_lookup <- tibble::tribble(
  ~LBTESTCD, ~PARAMCD, ~PARAM, ~PARAMN,
  "ALB", "ALB", "Albumin (g/L)", 1,
  "ALP", "ALKPH", "Alkaline Phosphatase (U/L)", 2,
  "AST", "AST", "Aspartate Aminotransferase (U/L)", 3,
  "CHOL", "CHOLES", "Cholesterol (mmol/L)", 4,
  "GGT", "GGT", "Gamma Glutamyl Transferase (U/L)", 5,
  "GLUC", "GLUC", "Glucose (mmol/L)", 6,
  "HBA1CHGB", "HBA1CHGB", "Hemoglobin A1C/Hemoglobin (mmol/mol)", 7,
  "INSULIN", "INSULIN", "Insulin (mIU/L)", 8,
  "TRIG", "TRIG", "Triglycerides (mg/dL)", 9
)
```

## Derive Core ADLB Variables {#derive_core}

The basic parameters and timing variables can be derived similarly to other BDS finding ADaMs.
For the derivation of Glucose and Insulin, only fasted results (where `LBFAST = "Y"`) are considered. For all other laboratory tests, the fasting status does not affect their inclusion in the dataset at this stage.

```{r}
# Define required ADSL variables for derivations
adsl_vars <- exprs(TRTSDT, TRTEDT, TRT01A, TRT01P)

adlb <- lb %>%
  # Remove non-fasted GLUC and INSULIN results
  filter(!(LBTESTCD %in% c("GLUC", "INSULIN") & LBFAST != "Y")) %>%
  # Join ADSL with LB (need TRTSDT for ADY derivation)
  derive_vars_merged(
    dataset_add = adsl,
    new_vars = adsl_vars,
    by_vars = get_admiral_option("subject_keys")
  ) %>%
  # Calculate ADT, ADY
  derive_vars_dt(
    new_vars_prefix = "A",
    dtc = LBDTC
  ) %>%
  derive_vars_dy(reference_date = TRTSDT, source_vars = exprs(ADT))

adlb <- adlb %>%
  # Add PARAMCD PARAM and PARAMN - from parameter lookup table
  derive_vars_merged_lookup(
    dataset_add = param_lookup,
    new_vars = exprs(PARAMCD, PARAM, PARAMN),
    by_vars = exprs(LBTESTCD)
  ) %>%
  ## Calculate PARCAT1 PARCAT2 AVAL AVALC ANRLO ANRHI
  slice_derivation(
    derivation = mutate,
    args = params(
      PARCAT1 = LBCAT,
    ),
    # Handle specific parameters requiring conventional units (CV)
    derivation_slice(
      filter = LBTESTCD %in% c("TRIG", "INSULIN"),
      args = params(
        PARCAT2 = "CV",
        AVAL = as.numeric(LBORRES),
        AVALC = NA_character_,
        ANRLO = as.numeric(LBORNRLO),
        ANRHI = as.numeric(LBORNRHI)
      )
    ),
    # Handle other parameters using standard units (SI)
    derivation_slice(
      filter = TRUE,
      args = params(
        PARCAT2 = "SI",
        AVAL = LBSTRESN,
        # Only populate AVALC if character value is non-redundant with AVAL
        AVALC = if_else(
          is.na(AVAL) | as.character(AVAL) != LBSTRESC,
          LBSTRESC,
          NA_character_
        ),
        ANRLO = LBSTNRLO,
        ANRHI = LBSTNRHI
      )
    )
  )
```

```{r echo=FALSE}
# Example display of derived data
dataset_vignette(
  arrange(adlb, USUBJID, PARCAT1, ADY, PARAMN),
  display_vars = exprs(!!!get_admiral_option("subject_keys"), ADT, ADY, PARAMCD, PARAM, PARAMN, PARCAT1, PARCAT2, AVAL, AVALC, ANRLO, ANRHI)
)
```

## Derive Visit Information {#derive_visits}

Derive Analysis Visit (`AVISIT`) and Analysis Visit Number (`AVISITN`).

```{r}
adlb <- adlb %>%
  mutate(
    AVISIT = case_when(
      str_detect(VISIT, "SCREEN") ~ "Baseline",
      !is.na(VISIT) ~ str_to_title(VISIT),
      TRUE ~ NA_character_
    ),
    AVISITN = case_when(
      AVISIT == "Baseline" ~ 0,
      str_detect(VISIT, "WEEK") ~ as.integer(str_extract(VISIT, "\\d+")),
      TRUE ~ NA_integer_
    )
  )
```

```{r echo=FALSE}
# Example display of derived data
dataset_vignette(
  arrange(adlb, USUBJID, PARCAT1, ADY, PARAMN),
  display_vars = exprs(!!!get_admiral_option("subject_keys"), PARAMCD, PARAM, PARCAT1, AVAL, ADY, AVISIT, AVISITN)
)
```

For deriving visits based on time-windows, see `{admiral}` [Visit and Period Variables](https://pharmaverse.github.io/admiral/articles/visits_periods.html).

## Derive Metabolic Parameters {#metabolic_params_desc}

In addition to the core ADLB variables, several metabolic parameters and scores will be derived based on laboratory data and vital signs. These derivations include:

*   **Fatty Liver Index (FLI)**: Derived from triglycerides (mg/dL), GGT (U/L), BMI (kg/m2), and waist circumference (cm).
 $$ FLI = \left[\frac{\exp(\lambda)}{1 + \exp(\lambda)}\right] \times 100 $$
where $$ \lambda = 0.953 \times \ln (\text{triglycerides}) + 0.139 \times BMI + 0.718 \times \ln (GGT) + 0.053 \times \text{waist circumference} – 15.745. $$
FLI is a score from 0 to 100. Screening risk bands on that score are derived below. A band is not a clinical diagnosis.

*   **Homeostasis Model Assessment – Insulin Resistance (HOMA-IR)**: Calculated from fasting plasma glucose (mmol/L) and fasting plasma insulin (mIU/L).
 $$ \text{HOMA-IR} = \frac{FPI \times FPG}{22.5}. $$
An alternative formula is used when fasting plasma glucose is assessed in mg/dL: $$ \text{HOMA-IR} = \frac{FPI \times FPG}{405}. $$

To derive these parameters and scores, the following variables will be needed:

*   From ADLB: TRIG (Triglycerides), GGT, GLUC (fasted), INSULIN (fasted).
*   From ADVS: BMI, WSTCIR (Waist Circumference).

### Add variables required for derivation {#merge}
In this section, we will merge relevant variables needed to derive metabolic parameter from the ADVS (Analysis Dataset for Vital Signs) dataset into the ADLB dataset. This is necessary to include variables such as BMI and Waist Circumference, which are required for the derivation of FLI score parameter in subsequent section.

```{r}
# Load ADVS dataset (assuming it has been created by ad_advs.R)
admiralmetabolic_advs <- admiralmetabolic::admiralmetabolic_advs
advs <- convert_blanks_to_na(admiralmetabolic_advs)

# Merge BMI and WSTCIR from ADVS to ADLB based on subject and date
adlb <- adlb %>%
  derive_vars_transposed(
    advs,
    by_vars = exprs(!!!get_admiral_option("subject_keys"), ADT),
    key_var = PARAMCD,
    value_var = AVAL,
    filter = PARAMCD %in% c("BMI", "WSTCIR")
  )
```

Please note that in this example we merge on ADT, but it is also possible to merge on AVISIT or other windowing algorithms specified in the statistical analysis plan.

```{r echo=FALSE}
# Example display of derived data
dataset_vignette(
  arrange(adlb, USUBJID, ADY, PARAMN),
  display_vars = exprs(!!!get_admiral_option("subject_keys"), PARAMCD, PARAM, AVAL, ADT, ADY, AVISIT, AVISITN, BMI, WSTCIR)
)
```

### Derive HOMA-IR index {#derive_homa}

This part describes how to calculate the Homeostasis Model Assessment – Insulin Resistance (HOMA-IR) index from fasting plasma glucose and fasting plasma insulin. This derivation does not require merging data from other datasets.

```{r}
# Derive HOMA-IR using derive_param_computed
adlb <- adlb %>%
  derive_param_computed(
    by_vars = exprs(!!!get_admiral_option("subject_keys"), AVISIT, AVISITN, ADT, ADY, !!!adsl_vars),
    parameters = c("INSULIN", "GLUC"),
    set_values_to = exprs(
      AVAL = AVAL.INSULIN * AVAL.GLUC / 22.5,
      PARAMCD = "HOMAIR",
      PARAM = "Homeostasis Model Assessment - Insulin Resistance",
      PARAMN = 10
    )
  )
```

```{r echo=FALSE}
# Example display of derived data
dataset_vignette(
  filter(adlb, PARAMCD == "HOMAIR") %>%
    arrange(USUBJID, ADY, PARAMN),
  display_vars = exprs(!!!get_admiral_option("subject_keys"), PARAMCD, PARAM, AVAL, ADY, AVISIT, AVISITN)
)
```

### Derive FLI score {#derive_fli}

To obtain the FLI score parameter, we'll utilize triglycerides, GGT, BMI, and waist circumference. Note that this derivation requires merging data from the ADVS dataset (BMI and Waist Circumference), as demonstrated in the previous section. The equation is unchanged. The next section only categorizes this score into screening risk bands.

```{r}
# Derive FLI using derive_param_computed
adlb <- adlb %>%
  derive_param_computed(
    by_vars = exprs(!!!get_admiral_option("subject_keys"), AVISIT, AVISITN, ADT, ADY, BMI, WSTCIR, !!!adsl_vars),
    parameters = c("TRIG", "GGT"),
    set_values_to = exprs(
      AVAL = {
        lambda <- 0.953 * log(AVAL.TRIG) + 0.139 * BMI + 0.718 * log(AVAL.GGT) + 0.053 * WSTCIR - 15.745
        (exp(lambda) / (1 + exp(lambda))) * 100
      },
      PARAMCD = "FLI",
      PARAM = "Fatty Liver Index",
      PARAMN = 11
    )
  )

adlb <- adlb %>%
  arrange(!!!get_admiral_option("subject_keys"), ADT, PARAMN) # Arrange for consistency
```

```{r echo=FALSE}
# Example display of derived data
dataset_vignette(
  filter(adlb, PARAMCD == "FLI") %>%
    arrange(USUBJID, ADY, PARAMN),
  display_vars = exprs(!!!get_admiral_option("subject_keys"), PARAMCD, PARAM, AVAL, ADY, AVISIT, AVISITN)
)
```

### Derive FLI screening risk bands {#fli_bands}

Categorize the existing FLI score with [`derive_vars_cat()`](https://pharmaverse.github.io/admiral/reference/derive_vars_cat.html). The three labels are screening risk bands only. They are not a clinical diagnosis.

The mid band includes 30 and excludes 60, so a score of exactly 60 is only in the upper band.

```{r}
# Screening risk bands on the existing FLI score. Not a diagnosis.
fli_avalcat_lookup <- exprs(
  ~PARAMCD, ~condition,             ~AVALCAT1,     ~AVALCA1N,
  "FLI",    AVAL < 30,              "<30",         1,
  "FLI",    AVAL >= 30 & AVAL < 60, "30–<60",      2,
  "FLI",    AVAL >= 60,             "≥60",         3,
  "FLI",    is.na(AVAL),            NA_character_, NA_integer_
)

adlb <- adlb %>%
  derive_vars_cat(
    definition = fli_avalcat_lookup,
    by_vars = exprs(PARAMCD)
  )
```

`BASECAT1` and `BASECA1N` keep the baseline screening risk band. The baseline record is the last non-missing result on or before treatment start, the same rule as BMI class in the ADVS template. Other parameters are not assigned an FLI band; their category variables stay missing.

```{r}
adlb <- adlb %>%
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(!!!get_admiral_option("subject_keys"), PARAMCD),
      order = exprs(ADT, AVISITN),
      new_var = ABLFL,
      mode = "last"
    ),
    filter = (!is.na(AVAL) & ADT <= TRTSDT)
  )

adlb <- adlb %>%
  derive_var_base(
    by_vars = exprs(!!!get_admiral_option("subject_keys"), PARAMCD),
    source_var = AVALCAT1,
    new_var = BASECAT1
  ) %>%
  derive_var_base(
    by_vars = exprs(!!!get_admiral_option("subject_keys"), PARAMCD),
    source_var = AVALCA1N,
    new_var = BASECA1N
  )
```

```{r echo=FALSE}
dataset_vignette(
  filter(adlb, PARAMCD == "FLI") %>%
    arrange(USUBJID, ADY, PARAMN),
  display_vars = exprs(
    !!!get_admiral_option("subject_keys"),
    PARAMCD, AVAL, AVALCAT1, AVALCA1N, ABLFL, BASECAT1, BASECA1N, AVISIT
  )
)
```

## Remaining ADLB Set-up {#adlb_end}

The `{admiral}` [Creating a BDS Finding ADaM vignette](https://pharmaverse.github.io/admiral/articles/bds_finding.html) describes further steps, including change from baseline, analysis flags (e.g., `ANL01FL`), reference ranges, and other common ADLB requirements. FLI screening risk bands, including the baseline band, are derived in the previous section.

# Example Scripts {#example}

ADaM | Sample Code
---- | --------------
ADLB | [ad_adlb.R](https://github.com/pharmaverse/admiralmetabolic/blob/main/inst/templates/ad_adlb.R){target="_blank"}
